---
title: "**Reflecting Landmarks**"
author: "Mark A. Conaway"
---
<br></br>

```{r, echo=F}
knitr::read_chunk("scripts/ReflectingLandmarks.R")
knitr::opts_chunk$set(dev.args=list(bg = 'transparent'), fig.align = 'center', class.source = ".custom-inline", class.output = ".custom-inline")
```
  
```{r prelims, include=FALSE}

```

## **Introduction**

When performing GM analyses with `geomorph` it is necessary that all landmark coordinates be aligned via [Generalized Procrustes Analysis](GPA) using the `gpagen` function. Of important note, however, is that the `gpagen` function does not automatically reflect specimens that are of a different side than the rest of a bilateral dataset (i.e. left vs. right).

Avoiding mismatched sides altogether is often not feasible due to the nature of data collection, however there is a quick way to reflect off-sided landmarks manually, which we will discuss here.


Let's start with a full dataset:

<div class = "code">
```{r, eval=FALSE, message=FALSE}
wings <- gpagen(mosquito$wingshape)

plotAllSpecimens(wings$coords)

```
</div>
<br></br>

```{r, plotdata, echo=FALSE, fig.align='center'}

```

Now here are the reflected specimens

<div class = "code">
```{r, eval=FALSE}
plotAllSpecimens(wings$coords*-1)
```
</div>
<br></br>

```{r, reflectspecimens, echo=FALSE}

```

A more likely scenario is that a user will need to reflect only a few specimens within their larger dataset. In order to do this, we must use what we learned in the tutorial on [3D Arrays](Arrays).

Here we have a dataset with one specimen (number four) that has been reflected.

<div class = "code">
```{r, eval=FALSE}
wings$coords[,,4] <- wings$coords[,,4]*-1
plotAllSpecimens(wings$coords)
```
</div>
<br></br>

```{r, onereflect, echo=FALSE}

```

We can reflect the specimen in question like so:

<div class = "code">
```{r, eval=FALSE}
wings$coords[,,4] <- wings$coords[,,4]*-1
plotAllSpecimens(wings$coords)
```
</div>
<br></br>

```{r, onereflect2, echo=FALSE}

```
<br></br>