---
title: "**Two-block Partial Least Squares**"
author: "Mark A. Conaway"
---

```{r, echo=F}
knitr::read_chunk("scripts/twobPLS.R")
knitr::opts_chunk$set(dev.args=list(bg = 'transparent'), fig.align = 'center', class.source = ".custom-inline", class.output = ".custom-inline")
```

```{r setup, warning=FALSE, include=FALSE}
library(geomorph)
data("plethShapeFood")
coords <- gpagen(plethShapeFood$land, print.progress = F)$coords
```

## **Introduction** 

A two-block partial least squares analysis is a useful tool for assessing the degree of association between two blocks (or sets) of Procrustes shape variables; or between one set of shape variables and other variables.

<div class = "func">
<details>
  <summary> <font size="+2"> `two.b.pls() (Expand for more details)` </font> </summary>

This function is used to perform a two-block partial least squares analysis in order to assess the degree if association between two blocks of variables.

* $A1$: A 3D array (p x k x n) containing Procrustes shape variables for the first block, or a matrix (n x variables)
* $A2$: A 3D array (p x k x n) containing Procrustes shape variables for the second block, or a matrix (n x variables)
* $iter$: Number of iterations for significance testing.The default is 999.
* $seed$: Optional argument for setting the seed for random permutations of the resampling procedure. If seed = "random," P-values for statistical comparisons will vary when the function is run multiple times on the same data. If left NULL (the default) P-values will be the same for repeated runs of the function, as long as the number of iterations (iter) remains the same.
* $print.progress$: A logical value (TRUE/FALSE) to indicate whether a progress bar should be printed to the screen. This is helpful for long-running analyses.
  
</details>
</div>
<br></br>

## **Example**

To perform this analysis we begin with our two blocks of variables. For this example we will use Procrustes-aligned coordinates and food use data from the `plethShapeFood` dataset available with geomorph. 

<div class = "code">
```{r, PLS}

```
</div>
<br></br>

Result of the PLS can be summarized like so:

<div class = "code">
```{r, sum}

```
</div>
<br></br>

For a simple visualization of the results, one can use the base R function `plot`:

<div class = "code">
```{r, eval=FALSE}
plot(PLS)
```
</div>
<br></br>

```{r, plot, echo=FALSE}

```

For more advanced visualizations, see [Visualizations]().

Finally, the `two.b.pls` function returns an object of class "pls" that is a list of the following objects, each of which can be accessed individually with the `$` operator:

* $r.pls$: The correlation coefficient between scores of the first singular vectors of the two blocks of variables.
* $P.value$ The empirically calculated P-value from the resampling procedure.
* $Effect.Size$: The multivariate effect size associated with sigma.d.ratio.
* $left.pls.vectors$: The singular vectors of the left (A1) block of variables.
* $right.pls.vectors$: The singular vectors of the right (A2) block of variables.
* $random.r$: The correlation coefficients found in each random permutation of the resampling procedure.
* $XScores$: Values of the left (A1) block projected onto singular vectors.
* $YScores$: Values of the right (A2) block projected onto singular vectors.
* $svd$: The singular value decomposition of the cross-covariances.
* $A1$: Input values for the left block.
* $A2$: Input values for the right block.
* $A1.matrix$: Left block (matrix) found from A1.
* $A2.matrix$: Right block (matrix) found from A2.
* $permutations$: The number of random permutations used in the resampling procedure.
* $call$: The full input code (match call).

<details>
  <summary> <font size="+2"> **Advanced Options for `two.b.pls`** </font> </summary>
* $Test$
</details>

<br></br>
