---
title: "**Calculating Interlandmark Distances**"
author: "Mark A. Conaway"
---
<br></br>

```{r, echo=F}
knitr::read_chunk("scripts/Interlmkdist.R")
knitr::opts_chunk$set(dev.args=list(bg = 'transparent'), fig.align = 'center', class.source = ".custom-inline", class.output = ".custom-inline")
```

```{r, prelims,eval=T,echo=F,results='hide',warning=FALSE,message=FALSE}

```



<div class = "func">
<details>
  <summary> <font size="+2"> `interlmkdist() (Expand for more details)` </font> </summary>
 
* $A$: A 3D [array]() containing landmark coordinates for a set of specimens.
* $lmks$: A matrix or dataframe of landmark numbers for the start and end points of each linear measurements. See below for specific formatting.

</details>
</div>
<br></br>

### Defining distances

The start and end points of the distance measure required need to be defined manually. This is done with a two-column matrix or dataframe with the column names "start" and "end." For the sake of this example, we are interested in the distances between landmarks 1:2 and 3:4 respectively:

<div class = "code">
```{r, defdists}

```
</div>
<br></br>

<div class = "code">
```{r, calcdists}

```
</div>

Each column contains one distance measure for each specimen. In this case, the first column is the distance between landmarks 1 and 2 for each of the 10 specimens in our dataset.
<br></br>

### Defining All Pairwise Distances

Calculating interlandmark distances for all possible combinations of landmarks requires a bit more code, but is nonetheless fairly simple:

First, we need to generate a vector of landmark numbers equal to the total number of landmarks in our dataset:

<div class = "code">
```{r, lmkvector}

```
</div>
<br></br>

Next, using the `combn` function we can generate a matrix of all possible combinations of those landmarks:

<div class = "code">
```{r, defalldists}

```
</div>
<br></br>

Then run the function as normal:
<div class = "code">
```{r, calcalldists}

```
</div>
