---
title: "**Identifying Outliers**"
author: "Dean Adams, Michael Collyer, Antigoni Kaliontzopoulou & Mark A. Conaway"
---
<br></br>

```{r, echo=F}
knitr::read_chunk("scripts/IdentifyOutliers.R")
knitr::opts_chunk$set(dev.args=list(bg = 'transparent'), fig.align = 'center', class.source = ".custom-inline", class.output = ".custom-inline")
```

```{r eval=TRUE, include=FALSE}
library(geomorph)
data(plethodon)
newland <- plethodon$land
newland[c(1,8),,2] <- newland[c(8,1),,2]
newland[c(3,11),,26] <- newland[c(11,3),,26]
```

## **Introduction**

All statistical analyses are sensitive to outliers. For instance, regression and most ordination approaches are heavily influenced by extremal data points, which can have a disproportionate leverage on the resulting statistical inferences. In GM data, one may have outliers for any number of reasons. First, specimens may be 'true' biological outliers; or there may be errors in digitizing (digitizing landmarks out of order is common); or any other number of reasons.

`Geomorph` provides a means of quickly identifying outliers for further inspection, using the function `plotOutliers`. Here we demonstrate its use, and provide a visualization of the shape of a potential outlier (relative to the mean shape), which aides in determining whether the outlier is something of concern. 

The function plots Procrustes distances ($\small{D}_{Proc}$) of all specimens relative to the mean. Large values are flagged as potential outliers, which can be subsequently inspected. Likewise, shape differences of outliers can be shown as well

<div class = "note">
**Important Note!** In order to use this function, one's data must be Procrustes aligned, using a Generalized Procrustes Analysis (GPA). There is a separate tutorial on aligning specimens using the geomorph function `gpagen` that can be found [here]().
</div>
<br></br>

<div class = "func">
<details>
  <summary> <font size="+2"> `plotOutliers() (Expand for more details)` </font> </summary>
  
* $A$: A [2D or 3D array](Arrays) containing Procrustes shape variables for a set of specimens
* $inspect.outliers$: A logical value (TRUE, or FALSE) indicating whether a plot(s) should be generated for outlying individuals comparing them to the consensus.

Note that for 3D specimens, the generated plot will be rendered in a separate window.

</details>
</div>
<br></br>

### **Input**

Now let's inspect for outliers in our `plethodon` salamander dataset


<div class = "code">
```{r eval=FALSE}
plotOutliers(Y$coords, inspect.outliers = T)
```
</div>
<br></br>


```{r, plotoutliers, eval=TRUE, echo=FALSE}

```
<br></br>

Here we have a plot of the shape distance of each specimen relative to the overall mean. Outliers (based on 95% CI) are shown in red. For this example we have two outliers. We can plot their shapes relative to the mean to reveal how their shape differs.
Note that for three-dimensional data, outlying specimens will be visualized in a separate, interactive window.

Notice that landmark plots of the outliers relative to the mean show the problem. Several landmarks were digitized out of order. We can now go back and correct our data prior to downstream analyses.</p>

<div class = "note">
**Note!** Because the [`gpagen`](Tutorials/2.1-GeneralizedProcrustesAnalysis.html) function does not automatically reflect landmark coordinates, sided elements that are a different side than the mean shape of a dataset (i.e. a left femur within a group of right femora), will be flagged as outliers.
</div>
<br></br>

# **Correcting Landmarking Errors in R**

The above example shows us that we have evidently placed some landmarks out of order. In particular, on specimen number 2, the coordinates for what should be landmark 8 are listed under landmark 1 and vice versa; the same is the case for landmarks 3 and 11 on specimen number 26.This can be corrected in R like so:

<div class = "code">
```{r}
newland[c(1,8),,2] <- newland[c(8,1),,2]
newland[c(3,11),,26] <- newland[c(11,3),,26]
Y <- gpagen(newland, print.progress = FALSE)
```
</div>
<br></br>

Now let's try plotting the outliers again:

<div class = "code">
```{r, eval=FALSE}
plotOutliers(Y$coords, inspect.outliers = T)
```
</div>
<br></br>

```{r, echo=FALSE}
plotOutliers(Y$coords, inspect.outliers = T)
```
<br></br>