---
title: "**Deleting Specimens and Landmarks**"
author: "Mark A. Conaway"
---
<br></br>

```{r, echo=F}
knitr::read_chunk("scripts/DeletingSpecLmk.R")
knitr::opts_chunk$set(dev.args=list(bg = 'transparent'), fig.align = 'center', class.source = ".custom-inline", class.output = ".custom-inline")
```

```{r prelims, include=FALSE}

```

## **Introduction**

On occasion, a researcher may need to remove one or more individuals or landmarks from their study. This could, for example, be because a specimen is damaged or doesn't fit the necessary criteria, or because a particular landmark turns out to be out-of-tolerance in terms of error.

The approach for deleting specimens and landmarks in R is similar to what we do for [Reflecting Landmarks](Link), in that we must access the level of interest in our [3D array](Link), and modify it. Likewise, the approach is slightly different depending on whether or not the specimens or landmarks to be deleted are sequential.

## **Deleting Specimens**

If we decide that a single specimen in our dataset should be removed, say specimen number four, we do the following:

<div class = "code">
```{r, rmsingle}

```
</div>
<br></br>

Alternatively, if multiple specimens *in sequence* need to be removed (say, 4 through 8), we can do this:

<div class = "code">
```{r, rmmulti}

```
</div>
<br></br>

One additional step is required to delete multiple specimens that are out-of-sequence. We must create a vector that lists the specimen numbers to be deleted, and then perform the above step with that vector, like so:

<div class = "code">
```{r, rmmultioutseq}

```
</div>
<br></br>

## **Deleting Landmarks**

The process for removing landmarks is essentially identical to removing specimens. We just need to access a different index of our [3D array](link).

Removing a single landmark:

<div class = "code">
```{r, rmlmk}

```
</div>
<br></br>

Removing multiple landmarks in-sequence:

<div class = "code">
```{r, rmmultilmk}

```
</div>
<br></br>

And multiple landmarks out-of-sequence:

<div class = "code">
```{r, rmmultilmkoutseq}

```
</div>
<br></br>
