---
title: "**Swapping Landmarks**"
author: "Mark A. Conaway"
---

```{r, echo=F}
knitr::read_chunk("scripts/SwappingLandmarks.R")
knitr::opts_chunk$set(dev.args=list(bg = 'transparent'), fig.align = 'center', class.source = ".custom-inline", class.output = ".custom-inline")
```

```{r, prelims, warning=FALSE, include=FALSE}

```

## **Introduction** 

One of the most common errors during the landmarking process is mis-ordered landmarks. Particularly when landmarks are close together it can be easy to place them in the wrong order. To correct this, we can swap the coordinates between two landmarks.
<br></br>

## **Swap two landmarks** 

Let's look at some example data and check for outliers:

<div class = "code">
```{r, eval=FALSE, results='hide'}
plotOutliers(gpagen(data)$coords)
```
</div>
<br></br>

```{r, plotoutliers, results='hide', echo=FALSE}

```

Extreme outliers like this are most often caused either by off-sided individuals who need to be [reflected](link), or in this case, by a landmarking error:

<div class = "code">
```{r, eval=FALSE}
plotRefToTarget(data[,,20], mshape(data), method = "vector", label = T)
```
</div>
<br></br>

```{r, plotRef, echo=FALSE}

```

Here, landmarks 1 and 7 have been placed in the wrong order on specimen number 20. To correct this, we simply:

<div class = "code">
```{r, eval=FALSE}
data[c(1,7),,20] <- data[c(7,1),,20]
plotRefToTarget(data[,,20], mshape(data), method = "vector")
```
</div>
<br></br>

```{r, swaptwo, echo=FALSE}

```

## **Swap multiple sets** 

In the event that multiple pairs of landmarks have been swapped on multiple specimens, the procedure is essentially the same, but requires more careful attention so as not to accidentally create a more severe error by swapping the wrong landmarks. Here we have a dataset with three specimens with two different landmarking errors:

<div class = "code">
```{r, eval=FALSE}
data <- gpagen(mosquito$wingshape)$coords
data[c(1,3,7,8),,c(4,8)] <- data[c(3,1,8,7),,c(4,8)]
data[c(1,9),,9] <- data[c(9,1),,9]
```
</div>
<br></br>


```{r, swapmulti, echo=FALSE, results='hide'}

```


```{r, lmkerrors, echo=FALSE, out.width= "200%"}

```

Here, landmarks 1 and 3, and 7 and 8 have been swapped on specimens 4 and 8. Additionally, landmarks 1 and 9 have been swapped on specimen 9. To correct this we need to run two separate lines of code:

<div class = "code">
```{r, swap}

```
</div>
<br></br>

And finally we check our work:
```{r, plot, echo=FALSE, out.width="200%"}

```
<br></br>

<div class = "note">
**Important Note!** This method is not a substitute for careful landmarking! If two landmarks are close enough to each other, but nevertheless swapped, that specimen may not register as an extreme outlier, or it may be less obvious that those landmarks have been swapped when plotted using `plotRefToTarget`.
</div>
<br></br>