---
title: "**Matching Specimens to a Phylogeny**"
author: "Mark A. Conaway"
---
<br></br>

```{r, echo=F}
knitr::read_chunk("scripts/MatchSpectoPhy.R")
knitr::opts_chunk$set(dev.args=list(bg = 'transparent'), fig.align = 'center', class.source = ".custom-inline", class.output = ".custom-inline")
```

```{r prelims, include=FALSE, warning=FALSE}

```


## **Introduction**
Occasionally when performing phylogenetic analyses, one will acquire phylogenetic data from a third-party source. In this case, the data represented in that phylogeny may or may not include the exact same species for which you have landmark data.

Here we describe a relatively easy way to compare and adjust phylogenetic trees to match input landmark data with taxonomic information, using the function `treedata` found in the R package `geiger`.

<div class = "func">
<details>
  <summary> <font size="+2"> `treedata() (Expand for more details)` </font> </summary>

* $phy$: A phylogenetic tree of class 'phylo'
* $data$: The input data. For landmark data, these must be in a format where all coordinates for a specimen are contained within a single row.
* $sort$: Logical value indicating whether to sort the data based on names found in `phy`.
* $warnings$: Logical value indicating whether the user should be notified of mismatched taxa.
  
</details>
</div>
<br></br> 

The data "plethspecies" embedded in `geomorph` includes both landmark data, and a phylogenetic tree. Crucially, one's landmark data must include species name information (see Array tutorial for more detail).

## **Using treedata**

The `treedata` function requires two components. First, an object of class "phylo," and second, landmark data with species name information. Importantly, for the `treedata` function to work, all landmark coordinates must be arranged by row. In other words, all landmark coordinates for a single specimen will be contained in a single row, with the row name being the species name, like so:

<div class = "code">
```{r, datbyrow}

```

</div>
<br></br>

Many functions in `geomorph` require one's data be in 3D array format. Likewise, the read.tps function imports landmark in 3D array format automatically. While this is useful for using geomorph, a 3D array cannot be used with the treedata function, because an array does not have row names in the same way that a simple matrix does.

Therefore, 3D arrays will need to be converted into a format that can be read by `treedata`. Luckily there is a function in geomorph that does this! Below we show the use of the `two.d.array` function to convert a 3D array of landmarks to a format that can be read by treedata:

<div class = "code">
```{r, twodarray}

```
</div>
<br></br>

After treedata has run, one can simply restore the landmark data to its previous format using the arrayspecs function. 

<br></br>