---
title: "GPA Introduction"
---

```{r, echo=F}
knitr::read_chunk("scripts/GPAIntro.R")
knitr::opts_chunk$set(dev.args=list(bg = 'transparent'), fig.align = 'center', class.source = ".custom-inline", class.output = ".custom-inline")
```

```{r, prelims, include=FALSE}

```

## **Introduction** 

Many morphometric analyses require that the researcher perform a Generalized Procrustes Analysis (GPA) in order to extract shape variables (that is, variables that do not include information on scale, translation, or rotation) from their landmark data. A GPA can be performed in `geomorph` using the function `gpagen`. This can be done with datasets composed of anatomical (fixed) landmarks as well as those with sliding semilandmarks.

<div class = "code">
```{r, eval=FALSE}
plotAllSpecimens(plethodon$land)
```
</div>
<br></br>

```{r, plotallspec, echo=FALSE}

```

<div class = "func">
<details>
  <summary> <font size="+2"> `gpagen() (Expand for more details)` </font> </summary>

* $A$: The input data to be aligned. Must be in [Array](2.1-AllAboutArrays.html) format.
* $curves$: An optional matrix defining which landmarks should be treated as curve semilandmarks. See below for more details.
* $surfaces$: An optional input defining which landmarks should be treated as surface semilandmarks.
* $print.progress$: Logical value indicating whether a progress bar should be printed to the screen.

This is a very large function with several other potential arguments which we describe in the [Advanced Options]() tutorial. See below for a full breakdown of the, also large, output of this function.

</details>
</div>
<br></br>

## **Output** 

The output of this function is a large list, or `gpagen` object with several subsets:

* $coords$: An array of Procrustes shape variables. In other words, the aligned landmark coordinates.
* $Csize$: A vector of centroid sizes for each specimen.
* $consensus$: The mean (consensus) configuration, or shape of the input data.
* $ProcD$: A Procrustes distance matrix for all specimens.
* $p$: The number of landmarks.
* $k$: The number of dimensions.
* $nsliders$: The number of curve semilandmarks.
* $nsurf$: The number of surface semilandmarks.
* $call$: The user-input code.

The following gpagen outputs are associated with arguments that are discussed elsewhere, in the [Advanced Options]() tutorial:

* $iter$: The number of GPA iterations until convergence was found.
* $points.VCV$: A variance-covariance matrix among Procrustes shape variables.
* $points.var$: Variances of landmark points.
* $data$: Data frame with a matrix of Procrustes shape variables and centroid size.
* $Q$: Final convergence criterion value.
* $slide.method$: Method used to slide semilandmarks.

Each of these subsets can be accessed separately using the '$' operator, like so:

