---
title: "**Pairwise Comparisons**"
author: "Mark A. Conaway"
---

```{r, echo=F}
knitr::read_chunk("scripts/Pairwise.R")
knitr::opts_chunk$set(dev.args=list(bg = 'transparent'), fig.align = 'center', class.source = ".custom-inline", class.output = ".custom-inline")
```

```{r setup, warning=FALSE, include=FALSE}
library(geomorph)
data("plethodon")
lmks <- gpagen(plethodon$land, print.progress = F)$coords
spec <- plethodon$species
site <- plethodon$site
data <- geomorph.data.frame(lmks, spec = spec, site = site)
```

## **Introduction** 

Whereas a [Procrustes ANOVA]() will tell the user the extent to which a variable X can be predicted by one or more variables Y-n, pairwise comparisons allow the user to quantify the statistical difference among different groupings of variables. To do this, we can use the RRPP function `pairwise`. This function allows the user to perform pairwise comparisons of group means of a linear model fit generated from a Procrustes ANOVA. 

Note that RRPP is a companion package which houses the underlying statistical permutational analytics for geomorph’s statistical procedures. It is installed and loaded simultaneously with geomorph.

<div class = "func">
<details>
  <summary> <font size="+2"> `pairwise()` </font> </summary>
 
* $fit$: A linear model fit. In our case, it is the object we created using `procD.lm`
* $groups$: A factor or vector that can be coerced into a factor that describes the groups for which we want to perform our pairwise comparisons with
* $print.progress$: A logical value (TRUE/FALSE) to indicate whether a progress bar should be printed on screen.
  
</details>
</div>
<br></br> 

## **Example Analysis**

For the purposes of this example, generating a model fit using the `procD.lm` function will be taken as-read. If you are not familiar with the `procD.lm` function please see the [ANOVA/Regression]() tutorial before proceeding here.

In addition to generating the fit, we will be adding material to our original [geomorph data frame](). Using the base R function `interaction`, we can create a factor with levels representing each pairwise permutation of our independent variables. In this case of this example, these are the "species," (Jord, Teyah) and "site," (Allo, Symp) variables.

<div class = "code">
```{r, fit}

```
</div>
<br></br>

After this preparation, we are ready to run the pairwise analysis:

<div class = "code">
```{r, pairs}

```
</div>
<br></br>

Like the `procD.lm` function, results of the `pairwise` function can be summarized like this:

<div class = "code">
```{r, sum}

```
</div>
<br></br>

Finally, the function generates an object of class "pairwise," which is a list containing the following:

<div class = "func">
<details>
  <summary> <font size="+2"> `pairwise` Output </font> </summary>

* $LS.means$: Least-squares (LS) means for groups, across permutations.
* $slopes$: Slopes for groups, across permutations.
* $means.dist$: Pairwise distances between means, across permutations.
* $means.vec.cor$: Pairwise vector correlations between means, across permutations.
* $means.lengths$: LS means vector lengths, by group, across permutations.
* $slopes.dist$: Pairwise distances between slopes (end-points), across permutations.
* $slopes.vec.cor$: Pairwise vector correlations between slope vectors, across permutations.
* $slopes.lengths$: Slope vector lengths, by group, across permutations.
* $slopes.diff.length$: Pairwise absolute differences between slope vector lengths, across permutations.
* $n$: Sample size.
* $p$: Number of variables
* $PermInfo$: List containing information for random permutations. This includes number of permutations (perms), the method used for permutation (perm.method), results of each permutation (perm.schedule), and the seed used for randomization (perm.seed).

</div>
<br></br>

<div class = "note">
**Important Note!** To see usage of this function incorporated with a full workflow, please see the [Group Comparisons](5.1-Group-Comparisons.html) tutorial.
</div>
<br></br>

<details>
  <summary> <font size="+2"> Advanced Options </font> </summary>

* $fit.null$: 
* $covariate$:

</details>
<br></br>
