---
title: "**Integration via Singular Warp Analysis**"
author: "Mark A. Conaway"
---

```{r, echo=F}
knitr::read_chunk("scripts/integrationtest.R")
knitr::opts_chunk$set(dev.args=list(bg = 'transparent'), fig.align = 'center', class.source = ".custom-inline", class.output = ".custom-inline")
```

```{r, prelims, warning=FALSE, include=FALSE}

```

## **Introduction** 

One potential avenue for quantifying the degree of morphological integration among sets of traits is to assess their covariance. This can be done with a two-block partial least squares analysis (PLS). However, in the context of integration studies, this is referred to as a Singular Warp Analysis. Similar to the [two.b.pls]() function, the `integration.test` function in geomorph allows the user to perform a PLS/Singular Warp analysis, but in this case for the specific purpose of assessing morphological integration among modular partitions of Procrustes coordinates. An added feature of `integration.test` is that the user may include more than two sets of variables for analysis.

<div class = "note">
**Important Note!**
When performing analyses with only two sets of variables, results will be identical to those returned by `two.b.pls`.
</div>
<br></br>



<div class = "func">
<details>
  <summary> <font size="+2"> `integration.test()` </font> </summary>
  
* $A1$: A [3D array (p x k x n)]() containing Procrustes shape variables for the first block, or a matrix (n x variables)
* $A2$: A 3D array (p x k x n) containing Procrustes shape variables for the second block, or a matrix (n x variables)
* $partition.gp$: A list of which landmarks (or variables) belong in which partition: (e.g. A, A, A, B, B, B, C, C, C). Required when only 1 dataset provided.
* $iter$: Number of iterations for significance testing
* $seed$: An optional argument for setting the seed for random permutations of the resampling procedure. If left NULL (the default), the exact same P-values will be found for repeated runs of the analysis (with the same number of iterations). If seed = "random", a random seed will be used, and P-values will vary. One can also specify an integer for specific seed values, which might be of interest for advanced users.
* $print.progress$: A logical (TRUE/FALSE) value to indicate whether a progress bar should be printed to the screen. This is helpful for long-running analyses.

</details>
</div>
<br></br>

## **Example of Analysis**

When using this function, one can input data in one of two ways. Firstly, the user can simply input a separate object for each block of data. Alternatively, however, the user can input a single object containing all Procrustes coordinates, and subsequently define, or partition, the separate blocks within that object using a vector. With two blocks of data, the function can be used similarly to `two.b.pls`: 

<div class = "code">
```{r, twopart}

```
</div>
<br></br>

Results can likewise be plotted using the base R plot function

<div class = "code">
```{r, eval=FALSE}
plot(IT)
```
</div>

```{r, twopartplot, echo=FALSE}
plot(IT)
```

Alternatively,  we can define a third partition in our input file:

```{r, threepart}

```

When we summarize the results, we get summary statistics for each pairwise combination of blocks:

```{r, threepartsum}
summary(IT2)
```

Note also that the initial r-PLS value is the average value for all blocks of data included in the analysis. 

<div class = "func">
<details>
  <summary> <font size="+2"> `integration.test` Output </font> </summary>

This function returns an object of class "pls" that contains the following values, each of which can be accessed with the $ operator.
  
*  $r.pls$:	The correlation coefficient between scores of projected values on the first singular vectors of left (x) and right (y) blocks of landmarks (or other variables). This value can only be negative if single variables are input, as it reduces to the Pearson correlation coefficient.
* $P.value$:	The empirically calculated P-value from the resampling procedure.
* $Effect.Size$: The multivariate effect size associated with sigma.d.ratio.
* $left.pls.vectors$:	The singular vectors of the left (x) block
* $right.pls.vectors$:	The singular vectors of the right (y) block
* $random.r$:	The correlation coefficients found in each random permutation of the resampling procedure.
* $XScores$:	Values of left (x) block projected onto singular vectors.
* $YScores$:	Values of right (y) block projected onto singular vectors.
* $svd$:	The singular value decomposition of the cross-covariances.
* $A1$:	Input values for the left block.
* $A2$:	Input values for the right block.
* $A1.matrix$:	Left block (matrix) found from A1.
* $A2.matrix$:Right block (matrix) found from A2.
* $permutations$:	The number of random permutations used in the resampling procedure.
* $call$:	The match call.
  
</details>
</div>
<br></br>  

<div class = "note">
**Important Note!**
If one wishes to incorporate a phylogeny in a PLS analysis, please see the tutorial for the [phylo.integration]() function.
</div>
<br></br>
