---
title: "**Comparison of Modularity**"
author: "Mark A. Conaway"
---

```{r, echo=F}
knitr::read_chunk("scripts/compareCR.R")
knitr::opts_chunk$set(dev.args=list(bg = 'transparent'), fig.align = 'center', class.source = ".custom-inline", class.output = ".custom-inline")
```

```{r,prelims, warning=FALSE, include=FALSE}

```

## **Introduction** 

When one is performing analyses of modularity on sets of traits in different samples, it is often of interest to determine whether those integration values are significantly different from one another. The function `compare.CR` allows one to statistically compare effect sizes of two or more PLS analyses; in particular, one might wish to compare modularity between two or more samples, each measuring modularity between separate sets of traits. Alternatively, this method can also be used to compare the degree of modular signal among alternative hypotheses of modularity for the same dataset.

This analysis calculates effect sizes as standard deviates, z, and performs two-sample z-tests, using the pooled standard error from the sampling distributions of the PLS analyses (Adams & Collyer, 2019).

The input for this function must be of class "pls," that is, an object resulting from the functions [modularity.test](), or [phylo.modularity](). Any number of objects can be input.

<div class = "func">
<details>
  <summary> <font size="+2"> `compare.CR()` </font> </summary>
  
* $...$: saved analyses of class CR
* $CR.null$: A logical (TRUE/FALSE) value to indicate whether a Null CR model (no modularity) should also be included in analysis. When comparing alternative hypotheses of modularity, this should be set to TRUE.
* $two.tailed$: A logical value to indicate whether a two-tailed test (typical and default) should be performed.

</details>
</div>
<br></br>

## **Comparing Modular Signal Across Datasets**

By way of example, we present an analysis of modularity between the body and operculum of pupfish. These data are included with `geomorph` by default, as a single object. The data have been separated by population (marsh vs. sinkhole) and sex for the purposes of this example. Please see the tutorials on  Data Manipulation and R Data Basic for information on how to accomplish this.

Here, the object 'group,' refers to the factor used to partition the pupfish data.

<div class = "code">
```{r, group}

```
</div>
<br></br>

Now, we run our modularity analyses with the help of the `map` function. This allows use to run the function on each of the partitioned sets of coordinates.

<div class = "code">
```{r, module}

```
</div>

This returns a list object containing the results of our four integration tests. Remember that objects within a list can be accessed using the $ operator.

Finally, we perform the statistical comparison of our results.

<div class = "code">
```{r, compare}

```
</div>

Summarizing these results, returns two tables of pairwise z-tests and p-values, as well as the effect sizes for each of the modularity analyses. 

## **Compare Alternative Modular Hypotheses**

Finally, we illustrate here how one might compare alternative hypotheses of modularity. For this example, we have modularity analyses on two separate partitions of the same data; one that hypotheses three modules (land.gps3), and one that hypotheses four modules (land.gps4). The process is similar to above. Note that we are running these analyses on female individuals only.

First, we run out modularity analyses:

<div class = "code">
```{r, altmod}

```
</div>

Then compare them. Note that, since we are comparing alternate hypotheses, the CR.null argument is set to TRUE.

<div class = "code">
```{r, altcomp}

```
</div>

The result of a summary in this case, returns tables of pairwise effect sizes and P-values. This also includes the null hypothesis of no modularity.

This function returns an object of class "compare.CR", which is a list containing the following:

<div class = "func">
<details>
  <summary> <font size="+2"> `compare.CR` output </font> </summary>
  
* $sample.z$:	A vector of effect sizes for each sample.
* $sample.r.sd$: A vector of standard deviations for each sampling distribution (following Box-Cox transformation).
* $pairwise.z$:	A matrix of pairwise, two-sample z scores between all pairs of effect sizes.
* $pairwise.p$:	A matrix of corresponding P-values.

</details>
</div>
<br></br>