---
title: "**Plot Shape Differences**"
author: "Mark A. Conaway"
---

```{r, echo=F}
knitr::read_chunk("scripts/plotreftarg.R")
knitr::opts_chunk$set(dev.args=list(bg = 'transparent'), fig.align = 'center', class.source = ".custom-inline", class.output = ".custom-inline")
```

```{r, prelims, warning=FALSE, include=FALSE}

```

## **Introduction** 

Functions like `plotAllSpecimens` allow the user to visualize shape variance within an entire set of specimens. If one wishes to visualize the shape differences between just two sets of coordinates - say, between the mean shape of a dataset, and another specimen of interest - one can use `plotRefToTarget`. 

<div class = "func">
<details>
  <summary> <font size="+2"> `plotRefToTarget()` </font> </summary>
  
  * $M1$: Matrix of landmark coordinates for the first (reference) specimen. A common choice here would be the mean shape of a set of specimens, accessed from a gpagen object with the operator `$consensus`.
  * $M2$: Matrix of landmark coordinates for the second (target) specimen. 
  * $mesh$: A mesh3d object for use with method="surface". See [read.ply]() for more information.
  * $outline$: An x,y curve or curves warped to the reference (2D only)
  * $method$: Method used to visualize shape difference (the type of plot); see below for details.
  * $mag$: The desired magnification to be used when visualizing the shape difference (e.g., mag=2).
  * $links$: An optional matrix defining for links between landmarks
  * $label$: A logical value (TRUE or FALSE) indicating whether landmark numbers will be plotted
  * $axes$: A logical value indicating whether the box and axes should be plotted (points and vector only)
  * $useRefPts$:

</details>
</div>
<br></br>

## **Method: TPS**

This method generates a thin-plate spline (TPS) deformation grid. Can be used for 2D or 3D data. For 3D data, deformations will be generated in both the xy and xz planes. For this, and each of the following examples, we compare the first specimen in our data set to the mean shape.

<div class = "code">
```{r, eval=FALSE}
plotRefToTarget(GPApleth$consensus, GPApleth$coords[,,1], links = links, method = "TPS")
```
</div>


```{r, tps, echo=FALSE}
plotRefToTarget(GPApleth$consensus, GPApleth$coords[,,1], links = links, method = "TPS")
```

## **Method: vector**

The vector show method, shows displacements in morphospace between corresponding landmarks in the reference (M1) and target (M2) specimen.

<div class = "code">
```{r, eval=FALSE}
plotRefToTarget(GPApleth$consensus, GPApleth$coords[,,1], links = links, method = "vector")
```
</div>


```{r, vec, echo=FALSE}
plotRefToTarget(GPApleth$consensus, GPApleth$coords[,,1], links = links, method = "vector")
```

## **Method: points**

This method plots landmarks of the target specimen **on top of** those of the reference. By default, the reference shape will be plotted in black.

<div class = "code">
```{r, eval=FALSE}
plotRefToTarget(GPApleth$consensus, GPApleth$coords[,,1], links = links, method = "points")
```
</div>


```{r, point, echo=FALSE}
plotRefToTarget(GPApleth$consensus, GPApleth$coords[,,1], links = links, method = "points")
```

## **Method: surface**

An input mesh3d object is warped using thin-plate spline. This illustrates shape differences between the reference and target specimen using a 3D object.

<div class = "code">
```{r, eval=FALSE}
plotRefToTarget(GPAscal$consensus, GPAscal$coords[,,1], mesh = mesh, method = "surface")

```
</div>
<br></br>

```{r, surf, echo=FALSE, results='hide'}
plotRefToTarget(GPAscal$consensus, GPAscal$coords[,,1], mesh = mesh, method = "surface")
```

```{r, echo=FALSE}
rglwidget()
```


