---
title: "**Plot 3D Specimen**"
author: "Mark A. Conaway"
---

```{r, echo=F}
knitr::read_chunk("scripts/plotspec.R")
knitr::opts_chunk$set(dev.args=list(bg = 'transparent'), fig.align = 'center', class.source = ".custom-inline", class.output = ".custom-inline")
```

```{r, prelims, warning=FALSE, include=FALSE}

```

## **Introduction** 

The geomorph function `plotspec` can be used to plot a 3D mesh of a specimen with an overlay of that specimen's landmarks.

<div class = "func">
<details>
  <summary> <font size="+2"> `plotspec()` </font> </summary>

* $spec$: An object of class [shape3d/mesh3d]()
* $digitspec$: Data matrix containing 3D fixed and/or surface sliding coordinates.
* $fixed$: The number of fixed template landmarks (see help files for `digitspec` function)
* $fixed.pt.col$: The color for plotting fixed template landmarks (if any)
* $fixed.pt.size$: Number representing the size for plotting fixed template landmarks (if any)
* $mesh.ptsize$: Numeric Size to plot the mesh points (vertices), e.g. 0.1 for dense meshes, 3 for sparse meshes
* $centered$: 	
Logical Whether the data matrix is in the surface mesh coordinate system (centered = FALSE) or if the data were collected after the mesh was centered (centered = TRUE)- see details.

</details>
</div>
<br></br>

<div class = "code">
```{r, eval=FALSE}
plotspec(spec = mesh, digitspec = lmks, fixed = 16, centered = T)
```
</div>
<br></br>

```{r, default, echo=FALSE}
plotspec(spec = mesh, digitspec = lmks, fixed = 16, centered = T)
#rglwidget()
```


The color of landmark points will default to red, but can be changed:

<div class = "code">
```{r, eval=FALSE}
plotspec(spec = mesh, digitspec = lmks, fixed = 16, centered = T, fixed.pt.col = "blue")

```
</div>


```{r, color, echo=FALSE}
plotspec(spec = mesh, digitspec = lmks, fixed = 16, centered = T, fixed.pt.col = "blue")
#rglwidget()
```

