---
title: "02c. Graphics"
author: "Dean Adams, Michael Collyer & Antigoni Kaliontzopoulou"
date: "3/16/2019"
output: html_document
---

## **Content description**
In geomorph, numerous functions allow for visualization of statistical results and of morphological shapes. Note that these functions are useful in conjunction with the analysis functions to provide visualizations of results. Below are several examples, but please consult the functon help pages for more details and examples on how to, e.g., define graphical parameters and other information.

Let´s first load some example data:
```{r eval=TRUE}
library(geomorph)

data(plethodon)
Y.gpa <- gpagen(plethodon$land, print.progress = F)    # GPA-alignment
```

## **1: Plot all specimens**
```{r eval=TRUE}
plotAllSpecimens(plethodon$land, links=plethodon$links)  # Raw data
plotAllSpecimens(Y.gpa$coords, links=plethodon$links)    # GPA-aligned data
```

## **2: PCA plot (tangent space)**
```{r eval=TRUE}
plotTangentSpace(Y.gpa$coords, groups = as.factor(paste(plethodon$species, plethodon$site)))
```

## **3: Visuallize shape differences between specimens**
Here we use the overall average as a reference, and deform it into some target, but this can be done with any two specimens you want to compare, e.g. group means, extremes of PC axes, minimum and maximum shape predicted across an allometry line etc.

```{r eval=TRUE}
ref <- mshape(Y.gpa$coords)
gps <- paste(plethodon$species, plethodon$site)
# Calculate group means
lsmeans <- arrayspecs((rowsum(two.d.array(Y.gpa$coords), gps)/as.vector(table(gps))), 12, 2) 

plotRefToTarget(ref,lsmeans[,,1], links=plethodon$links)
plotRefToTarget(ref,lsmeans[,,1], links=plethodon$links, mag=2)  #magnified 2X
plotRefToTarget(ref,lsmeans[,,1], links=plethodon$links, mag=5,
                method="vector", gridPars=gridPar(pt.bg="red", link.col="green"))
```

## **4: Outlier plot**
The function plotOutliers returns dimnames and address of all specimens ordered by their Procrustes distance from the mean shape. One can also ask for visualization of the outlier shapes as compared to the mean.
```{r eval=TRUE}
# Let's first make some outliers
newland <- plethodon$land
newland[c(1,8),,2] <- newland[c(8,1),,2]
newland[c(3,11),,26] <- newland[c(11,3),,2]
Y <- gpagen(newland, print.progress = FALSE) 

plotOutliers(Y$coords, inspect.outliers = T)
``` 

## **5: Allometry plot**
```{r eval=TRUE}
data(ratland)
Y.gpa.rat <- gpagen(ratland, print.progress = FALSE)
gdf <- geomorph.data.frame(shape = Y.gpa.rat$coords, size = Y.gpa.rat$Csize)
rat.allom <- procD.lm(shape~size, data=gdf, print.progress = FALSE)
summary(rat.allom)
plot(rat.allom, type = "regression", reg.type = "CRC", predictor = Y.gpa.rat$Csize) 
plot(rat.allom, type = "regression", reg.type = "RegScore", predictor = Y.gpa.rat$Csize)
plot(rat.allom, type = "regression", reg.type = "PredLine", predictor = Y.gpa.rat$Csize)
```

## **6: PhyloMorphospace plot**
```{r eval=TRUE}
data(plethspecies)
Y.gpa <- gpagen(plethspecies$land, print.progress = FALSE)    
plotGMPhyloMorphoSpace(plethspecies$phy, Y.gpa$coords)
plotGMPhyloMorphoSpace(plethspecies$phy, Y.gpa$coords, zaxis='time')
```

## **7: 3D warping**
```{r eval=TRUE}
### DEAN: DOESN`T WORK NOT SURE WHY
scallops <- readland.tps("Data/scallops for viz.tps", specID = "ID")
ref <- mshape(scallops)
refmesh <- warpRefMesh(read.ply("Data/glyp02L.ply"), 
                       scallops[,,1], ref, color=NULL, centered=T)
plotTangentSpace(scallops, axis1 = 1, axis2 = 2, warpgrids=T, mesh= refmesh)
```

