---
title: "04b. TrajectoryAnalysis"
author: "Dean Adams, Michael Collyer & Antigoni Kaliontzopoulou"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

## **Content description**
In geomorph, there are two approaches for estimating missing data:

* First, one can use the thin-plate spline to warp the landmark locations in the reference to those in the target specimen, and use the locations to represent missing landmarks. 
* Second, one can perform a multivariate regression (implemented using PLS) to perform the estimation.

Load the package and example data, and then creat some missing landmarks.
```{r eval=TRUE}
library(geomorph)
data(plethodon)

# Create missing landmarks
plethland <- plethodon$land
plethland[3,,2]<-plethland[8,,2]<-NA  
plethland[3,,5]<-plethland[8,,5]<-plethland[9,,5]<-NA
plethland[3,,10]<-NA
plethland[,,1:3]
```

## **TPS landmark estimation**
```{r eval=TRUE}
noNA.tps <- estimate.missing(plethland, method="TPS")
noNA.tps[,,1:3]
```

## **Regression landmark estimation**
```{r eval=TRUE}
noNA.reg <- estimate.missing(plethland,method="Reg")
noNA.reg[,,1:3]
```

## **Compare results** 
Here we know the truth (original landmark coordinates before we eliminated them) so we can compare the performance of different methods
```{r eval=TRUE}
res <- gpagen(plethodon$land, print.progress = FALSE)
res.tps <- gpagen(noNA.tps, print.progress = FALSE)
res.reg <- gpagen(noNA.reg, print.progress = FALSE)

pls.ori.tps <- two.b.pls(res$coords, res.tps$coords, print.progress = FALSE)
plot(pls.ori.tps)
summary(pls.ori.tps)

pls.ori.reg <- two.b.pls(res$coords, res.reg$coords, print.progress = FALSE)
plot(pls.ori.reg)
summary(pls.ori.reg)

pls.tps.reg <- two.b.pls(res.tps$coords, res.reg$coords, print.progress = FALSE)
plot(pls.tps.reg)
summary(pls.tps.reg)


```

