---
title: "01.1: Digitizing Landmarks in StereoMorph"
author: ""
output: html_document
---

## **Content Description**

This tutorial summarizes the basic steps of digitizing 2D GM data using the R-package StereoMorph, and then importing them to `geomorph`. As `StereoMorph` uses Shiny to allow the user to digitize, it is interactive, so you need to run the code to gain hands-on experience on the details of how digitizing proceeds.

You will need to install and load `StereoMorph` first.
```{r eval=FALSE}
install.packages("StereoMorph")

```

```{r eval=TRUE}
library(StereoMorph)
library(geomorph)

```


## **Digitizing Landmarks**

First, let´s see how one can digitize fixed landmarks. These correspond to well-defined anatomical points on your structure of interest. You will need a folder with the pictures you will digitize (Images, in our example)
In this script, digitized data are stored in another folder, called Shapes
To digitize, e.g., five fixed landmarks, run: 
```{r eval=FALSE}
digitizeImages(image.file='Data/Fish-Images', shapes.file='Data/Fish-Shapes-ex',
               landmarks.ref=paste("LM", c(1:5), sep=""))
```

To digitize for the first time:

* Go to Landmarks tab to see list of landmarks
* The first lm is in bold
* In the image window, double-click on the desired anatomical location to digitize
* The point is now selected, and can be edited, moved etc. (is highlighted in green)
* Select the next lm name and repeat
* Navigate between landmarks with "n" (next) and "p" (previous)
* Select lm and press "d" to delete
* Save before moving to the next picture (or select the automatic save option)

* NOTE: one has the ability to copy all landmarks to each image, and then move them to the desired locations. This can be useful so that landmarks remain in the correct order.

---

#### **IMPORTANT DETAIL: MISSING LANDMARKS!!** Any specimen with missing data (e.g., a broken specimen) may still be digitized. In this case, simply leave the missing landmarks **BLANK**. They will be treated as `NA` and can be estimated in subsequent data-processing steps.

#### **IMPORTANT DETAIL: DELETING LANDMARKS!!** To delete a landmark, click on the lm, and hit 'D'.

---

## **Scaling**

To scale landmark coordinates during superimposition and obtain accurate estimates of the relative size of the studied objects, you need to define the scaling of the images used, i.e. to indicate the correspondence of picture pixels with a "true" size measurement. For this: 

* Go to the scaling tab
* Select the ruler points by double clicking (one can use >1 scaling intervals, and then an average is calculated)
* Introduce your reference interval and units
* Save 

## **Digitizing Curves**

The easiest way to obtain sliding semilandmarks to describe the shape of curves, is by first digitizing a curve in StereoMorph and then sampling the desired number of points from it. Curves are digitized in StereoMorph using a Bezier approximation. For this you need to: 

1. Specify curve names and start/end landmarks (see example_curve.txt)
*NOTE:* do not use spaces or strings used to name landmarks when naming the curves
2. Re-open the digitizer to record the curves

```{r eval=FALSE}
digitizeImages(image.file='Data/Fish-Images', shapes.file='Data/Fish-Shapes-ex',
               landmarks.ref=paste("LM", c(1:5), sep=""),
               curves.ref = "Data/example.curves.txt")
```

You can then:

* Add points to the curve by selecting an empty line (with "-") and double clicking on the image
* After the first point, you need to add control points two at a time, to keep the curve continuous
* Navigate across points with "n" and "p" and move them to adjust the curve

Once the curve is digitized, you can **subsample a desired number of semilandmarks** from it.  This is accomplished easiest with the function, **readland.shapes**.  This is a geomorph function that reads in data from a class `shapes` object.  (In geomorph, there are several readland.___ functions for different landmark file types.)  Before using `readland.shapes`, a shapes object must be created (a Stereomorph function).
```{r eval=TRUE}

shapes <- readShapes("Data/Fish-Shapes-ex")
shapesGM <- readland.shapes(shapes, nCurvePts = 10) # can change 10 to a different number

# Perform GPA
Y.gpa <- gpagen(shapesGM, print.progress = FALSE)
plot(Y.gpa)

```

When one has completed a serious digitizing effort, the `readland.shapes` function offers extreme flexibility for choosing the sets of landmarks to use as a definition of "shape".  **It must be stated that for those of us who have spent some years in GMM, this is profoundly amazing!**  Not too long ago, one must have had an *a priori* definition of a landmark configuration and digitize every specimen according to a preformed plan.  Now, one can experiment and find an optimal landmark configuration, post-digitizing.

Too demonstrate this, let's use a digitizing set with 11 landmarks and 9 curves.  First we need to define how many landmarks should be in each curve.

```{r eval=TRUE}

shapes <- readShapes("Data/example.digitized")

# curves include eye, head, tail 1, tail end, tail 2, anal fin, 
# pectoral fin, opercle, pre-opercle
curves.list <- c(20, 24, 24, 16, 12, 8, 8, 24, 16)

```

Then use `readland.shapes` to read the data into a useable format for **geomorph**.

```{r eval=TRUE}

shapesGM <- readland.shapes(shapes, nCurvePts = curves.list)

```

Every curve needs to have at least three landmarks (or it is not a curve).  Other than that, one can vary the number for every curve.  The set chosen is hypothetical.  Let's see what it looks like following generalized Procrustes Analysis (GPA).

```{r eval=TRUE}

Y.gpa <- gpagen(shapesGM, print.progress = FALSE)
plot(Y.gpa)

```

The density of semilandmarks might be a little too great.  We can reduce it by 50%, for example, in one line of code:

```{r eval=TRUE}

shapesGM <- readland.shapes(shapes, nCurvePts = 0.5 * curves.list)
Y.gpa <- gpagen(shapesGM, print.progress = FALSE)
plot(Y.gpa)

```

Later it might impress you that no arguments for gpagen are needed when reading data in with `readland.shapes`.  Class `geomorphShapes` objects also have various attributes that can be exported for other uses.

```{r eval=TRUE}
class(shapesGM)

attributes(shapesGM)

shapesGM$fixed # which landmakrs are fixed
shapesGM$curves # a curves matrix for gpagen - can be modified
shapesGM$landmarks[[1]] # specimen 1
names(shapesGM$landmarks) # all specimen names
```

