---
title: "01.4: GPA: Generalized Procrustes Analysis"
author: ""
output: html_document
---

## **Content Description**

Generalized Procrustes Analysis (GPA) aligns our landmark configurations to a common coordinate system, and allows us to extract shape (and size) data from them. The function `gpagen` performs this procedure. Here we demonstrate its use for landmark data: 

## **Performing GPA Superimposition**

```{r eval=TRUE}
library(geomorph)
data(plethodon)
pleth.gpa <- gpagen(plethodon$land, print.progress = F)
summary(pleth.gpa)
```

Using `summary` provides us with a general description of the analysis we performed. Additionally, there are quite a few components to our GPA object: the most important of which are `$coords` and `$Csize`$

```{r eval=TRUE}
attributes(pleth.gpa)
```


## **Plotting GPA-Aligned Specimens**

```{r eval=TRUE}
plot(pleth.gpa)
plotAllSpecimens(pleth.gpa$coords, links = plethodon$links)
```

Use of `plot.param` can facilitate additional flexibility (change point or link colors, thickness, etc.)
