---
title: "01.4: GPA: Generalized Procrustes Analysis"
author: "Dean Adams, Michael Collyer & Antigoni Kaliontzopoulou"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

## **Content Description**

As we have seen in the theory classes, Generalized Procrustes Analysis is the main tool that allows one to extract shape (and size) variables from landmark coordinates . This procedure standardizes for size, location and rotation effects and provides meaningful shape data for downstream statistical analyses.  Let's take a look at how one performs GPA in `geomorph`, using the function `gpagen`:

## **Performing GPA Superimposition**

```{r eval=TRUE}
library(geomorph)
data(plethodon)
pleth.gpa <- gpagen(plethodon$land, print.progress = F)
summary(pleth.gpa)
```

Using `summary` provides us with a general description of the analysis we performed. Additionally, there are quite a few components to our GPA object: the most important of which are `$coords` and `$Csize`$

```{r eval=TRUE}
attributes(pleth.gpa)
```


## **Plotting GPA-Aligned Specimens**

Of course, one wishes to plot the shapes once they are superimposed. We can do this with one of several functions, shown here:

```{r eval=TRUE}
plot(pleth.gpa)
plotAllSpecimens(pleth.gpa$coords, links = plethodon$links)
```

Use of `plot.param` can facilitate additional flexibility (change point or link colors, thickness, etc.)

### Note!

As we learned incidentally, class `geomorphShapes` objects can be loaded directly into `gpagen`.  Whereas most data read into `geomorph` produce a 3D array, only, the landmarks are merely one component of a class `geomorphShapes` object.  They are also a list rather than a 3D array.  The `gpagen` function can work with this list but other functions, like `plotAllSpecimens` require an array.  There is a base R function, `simplify2array`, which can convert a list to an array, making landmarks useable in other functions; e.g., 

```{r eval=TRUE}
library(StereoMorph)
shapes <- readShapes("example.digitized")
shapesGM <- readland.shapes(shapes) # just the fixed landmarks
justLM <- simplify2array(shapesGM$landmarks)
plotAllSpecimens(justLM)
plot(gpagen(justLM, print.progress = FALSE))
```




