---
title: "02: Day 2 Material"
author: ""
output: html_document
---

```{r setup, include=FALSE, echo = TRUE, tidy = TRUE}
library(knitr)
library(geomorph)
opts_chunk$set(echo = TRUE)
```

## General Introduction

Today we will briefly introduce some `geomorph` functions for performing the analyses described yesterday. **Note:** You should be able to read in your data and perform the analyses from yesterday! Thus this tutorial augments the 'next steps' of a morphometric analysis (i.e., we are beginning to flesh out the steps of the Procrustes Paradigm for geometric morphometric analyses).

## 1: Generalized Procrustes Analysis (GPA) with Sliding Semilandmarks

Below is a simple 2D example containing points and curves. For 3D we could add curves and/or surfaces (see help file)

```{r eval=TRUE}
data(hummingbirds)
hummingbirds$curvepts   

gpa.BE <- gpagen(hummingbirds$land, curves=hummingbirds$curvepts, ProcD=FALSE, print.progress = F)
plot(gpa.BE)

gpa.procD <- gpagen(hummingbirds$land, curves=hummingbirds$curvepts, ProcD=TRUE, print.progress = F)
plot(gpa.procD)
```

##### NOTE: For StereoMorph users, the notation is slightly different for specifying curves (see help file):

```{r eval=TRUE}
shapes <- StereoMorph::readShapes("Data/example.digitized")
shapesGM <- readland.shapes(shapes, 
  nCurvePts = c(12, 12, 12, 8, 6, 6, 6, 12, 10))

Y.gpa <- gpagen(shapesGM, print.progress = FALSE)
plot(Y.gpa)
```

## 2: Partial Least Squares (PLS)

Here is a simple example of partial least squares: 

```{r eval=TRUE}
data("plethShapeFood")
Y.gpa <- gpagen(plethShapeFood$land,
                print.progress = FALSE,)
food <- plethShapeFood$food
rownames(food) <- names(Y.gpa$Csize) # just to assure match

PLSfood <- two.b.pls(food, Y.gpa$coords, 
                     print.progress = FALSE,
                     iter = 9999)
plot(PLSfood)

```

## 3: GLM: Regression

Here is a simple example of regression from a linear model:

```{r eval=TRUE}
fit.null <- procD.lm(coords ~ 1, data = Y.gpa,
                     iter = 9999)
fit.alt <- procD.lm(coords ~ log(Csize), data = Y.gpa,
                    turbo = FALSE, iter = 9999)
anova(fit.alt)
plot(fit.alt, type = "regression", reg.type = "RegScore", predictor = Y.gpa$Csize)
```

## 4: Allometry

Here is a simple example of examining allometry: 

```{r eval=TRUE}
data(pupfish)
plotAllSpecimens(pupfish$coords)  #NOTE: already GPA-aligned
  #Y.gpa <- gpagen(pupfish$coords, print.progress = FALSE)    #GPA-alignment  
pupfish$logSize <- log(pupfish$CS)
pupfish$Group <- interaction(pupfish$Pop, pupfish$Sex)

fit <- procD.lm(coords ~ logSize, data = pupfish, print.progress = FALSE) 
anova(fit)

plot(fit, type = "regression", reg.type = "RegScore", predictor = pupfish$logSize, pch = 19)
  #NOTE: Try also: reg.type = "PredLine"
```
   
#### Comparing Allometric Trajectories

```{r eval=TRUE}
fit.common <- procD.lm(coords ~ logSize + Group, 
                           data = pupfish, print.progress = FALSE) 
fit.unique <- procD.lm(coords ~ logSize * Group, 
                           data = pupfish, print.progress = FALSE)  
anova(fit.unique)

par(mfcol = c(1,2))
plot(fit.common, type = "regression", predictor = pupfish$logSize, 
     reg.type = "PredLine", pch=19, col = pupfish$Group)
legend("topleft", legend = unique(pupfish$Group), pch = 21, pt.bg = unique(pupfish$Group))
mtext("Common Slopes")
plot(fit.unique, type = "regression", predictor = pupfish$logSize, 
     reg.type = "PredLine", pch=19, col = pupfish$Group)
legend("topleft", legend = unique(pupfish$Group), pch = 21, pt.bg = unique(pupfish$Group))
mtext("Unique Slopes")
par(mfcol = c(1,1))
```

**Conclusion:** Not much going on for this example (**illustration purposes only**). 

## 5: Things to Explore on Your Own

1. Remember Day 1 (read in your data, GPA, PCa, TPS)
2. Perform GPA + sliders
3. 3D data: Perform GPA with points, curves, surfaces, or some combination
4. Perform PLS: visualize shapes along PLS
5. Perform regression: visualize shapes along regression
6. Generate shapes for small and large specimens in an allometry analysis 
7. **Explore function options and output** (there are many!)

