---
title: "02.3: Regression and covariation"
author: "Dean Adams, Michael Collyer & Antigoni Kaliontzopoulou"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

## **Content Description**

In lecture, we discussed several ways of evaluating covariation between components of our dataset. One approach: partial least squares (PLS), evaluates the degree of covariation between two blocks of variables. The other method, regression, uses a linear model framework for evaluating covariation. 

In `geomorph`, both PLS and regression models are possible. Here we demonstrate both: 

## **1: 2B-PLS**

One common covaration analysis is partial least squares (PLS), which evaluates the degree of covariation between two sets of variables. The function `two.b.pls` in `geomorph` faciliates this analysis.

First, let's load some data: here we will use the pupfish dataset, where heads and bodies are considered two blocks of variables. 

```{r eval=TRUE}
library(geomorph)
data(pupfish)
Y.gpa <- gpagen(pupfish$coords, print.progress = FALSE)
plot(Y.gpa)

```

Now let's evaluate covariation between head shape and body shape

```{r eval=TRUE}
shape <- Y.gpa$coords
headland <- c(4, 10:17, 39:56)

PLS <- two.b.pls(shape[headland,,],shape[-headland,,], iter=999, print.progress = FALSE)
summary(PLS)
plot(PLS)

```

The two are quite tightly associated. Finally, let's visualize shape trends along this PLS axis. For this we use `shape.predictor`, as shown earlier for PCA and group mean predictions: 

```{r eval=TRUE}
preds <- shape.predictor(shape[headland,,], two.d.array(shape[-headland,,]), Intercept = FALSE,
                         method = "PLS", pred1 = -0.2, pred2 = 0.2) # using PLS plot as a guide

M <- mshape(shape[headland,,])
plotRefToTarget(M, -1*preds$pred1, mag=3)
plotRefToTarget(M, preds$pred2, mag=3)
```

These represent the changes in head shape across the PLS association axis (similar plots could be generated for body shape).

## **2: Regression**

The second analytical approach for evaluating multivariate associations is regression. In `geomorph`, regression is accomplished using `procD.lm`. This function has considerable options and flexibility, some of which are shown here (*MANY* others are demonstrated in the ANOVA lecture). Here we fit a regression of `shape~log(Csize)`. 

```{r eval=TRUE}
pupfish$logSize <- log(pupfish$CS)  #add logCS to geomorph data frame
fit <- procD.lm(coords ~ logSize, data = pupfish, print.progress = FALSE)
anova(fit)
```

For multivariate regression, we have several plotting options available. One contains diagnostic plots, which are useful in evaluating the robustness of the regression fit. The others are visualization plots: via predicted lines (Adams and Nistri 2010), and Regression Scores (Drake and Klingenberg 2008).

```{r eval=TRUE}
plot(fit)
plot(fit, type = "regression", reg.type = "PredLine", predictor = pupfish$logSize, pch=21, bg="red")
plot(fit, type = "regression", reg.type = "RegScore", predictor = pupfish$logSize, pch=21, bg="red")
```

Finally, we can visualize shape changes along the regression using `shape.predictor`. These represent predicted shapes along the regression axis:

```{r eval=TRUE}
allom.plot <- plot(fit, 
                   type = "regression", 
                   predictor = pupfish$logSize,
                   reg.type ="PredLine") # make sure to have a predictor 

preds <- shape.predictor(allom.plot$GM$fitted, x= allom.plot$RegScore, Intercept = FALSE, 
                         predmin = min(allom.plot$RegScore), 
                         predmax = max(allom.plot$RegScore)) 
M <- mshape(pupfish$coords)
plotRefToTarget(M, preds$predmin, mag=3)
plotRefToTarget(M, preds$predmax, mag=3)
```

### Notes

1. There is also a plot function, specifically for generating allometry plots, `plotAllometry`.  It is no different than `procD.lm`, `type = "regression"` when the predictor is a size variable, but it might seem more intuitive for some users.  For example:

```{r eval=TRUE, fig.height=4, fig.width=8}
par(mfcol = c(1,2))
plot(fit, 
     type = "regression", 
     predictor = pupfish$logSize,
     reg.type ="RegScore") 

mtext("plot.procD.lm")

plotAllometry(fit,
              size = pupfish$CS,
              logsz = TRUE,
              method = "RegScore")
mtext("plotAllometry")
```

```{r echo=FALSE}
par(mfcol = c(1,1))
```

2. The `picknplot.shape` function allows interactive shape prediction in any of the plots generated here.  As it is interactive, one would have to run the script associated with this lab tutorial to get a feel for how it works.