---
title: "03.1: Allometry"
author: ""
output: html_document
---

## **Content Description**

Allometry, is the covariation of shape with size, and is a critically-important pattern to evaluate in morphometric analyses. Here we show several analyses available in `geomorph` for investigating shape-size relationships. 

## **1: Allometry: A Simple Example**

```{r eval=TRUE}
library(geomorph)
data(pupfish)
plotAllSpecimens(pupfish$coords)  #NOTE: already GPA-aligned
  #Y.gpa <- gpagen(pupfish$coords, print.progress = FALSE)    #GPA-alignment  
pupfish$logSize <- log(pupfish$CS)
pupfish$Group <- interaction(pupfish$Pop, pupfish$Sex)

fit <- procD.lm(coords ~ logSize, data = pupfish, print.progress = FALSE) 
anova(fit)

# Predline (note: `plotAllometry` can also be used)
plot(fit, type = "regression", reg.type = "PredLine", predictor = pupfish$logSize, pch = 19)

# RegScore (note: `plotAllometry` can also be used)
plot(fit, type = "regression", reg.type = "RegScore", predictor = pupfish$logSize, pch = 19)

# CAC
plotAllometry(fit, size = pupfish$logSize, logsz = FALSE, method = "CAC", pch = 19)
```

# Now we plot predicted shapes (see day 3 script)

```{r eval = T}
M <- mshape(pupfish$coords)
preds <- shape.predictor(fit$GM$fitted, x= pupfish$logSize, Intercept = TRUE, 
                         predmin = min(pupfish$logSize), 
                         predmax = max(pupfish$logSize)) 
par(mfrow=c(1,2))
plotRefToTarget(M, preds$predmin, mag=1)
mtext("Regression Min")
plotRefToTarget(M, preds$predmax, mag=1)
mtext("Regression Max")
par(mfrow=c(1,1))
```



## **2: Group Allometries**

Next, we add some biological realism, and consider the case of allometry patterns in multiple groups. This, of course, represents an ANCOVA design: analysis of covariance. It requires a more complex linear model, and affords additional possibilities in terms of analytical approaches, and visualizations.

#### Note!  Before ever launching into an investigation of allometric differences, it is wise to ask if groups have a common allometry or unique allometrices.

```{r eval=TRUE}
fit.common <- procD.lm(coords ~ logSize + Group, 
                           data = pupfish, print.progress = FALSE) 
fit.unique <- procD.lm(coords ~ logSize * Group, 
                           data = pupfish, print.progress = FALSE)  
anova(fit.unique)
```

In this case, the unique slopes are **not** an improvement of fit. This suggest that the slopes are relatively parallel. However, but for the purposes of illustration we will continue as if they were. 

## **2a: Pairwise Comparisons**

Based on the outcome of a homogeneity of slopes test, one may wish to perform different types of pairwise comparisons. These come in the form of: 1) comparisons of adjusted least-squares means (common slopes), 2) comparisons of slope angles (unique slopes), 3) comparisons of regression lengths (unique slopes), and 4) comparisons of group disparity. For the present example, #2 and #3 are most appropriate to investigate.

```{r eval=TRUE}
slope.pw <- pairwise(fit.unique, fit.null = fit.common, 
                     groups = pupfish$Group,
                     covariate = pupfish$logSize, print.progress = FALSE)
summary(slope.pw, test.type = "VC", angle.type = "deg") # angular differences
summary(slope.pw, test.type = "dist", angle.type = "deg") # amount of shape change differences
```

**Conclusion:** Not much going on for this example (**illustration purposes only**). 

## **2b: Pairwise Comparisons**

This shows visually that the common slopes model is preferred.

```{r eval=TRUE}
par(mfcol = c(1,2))
plot(fit.common, type = "regression", predictor = pupfish$logSize, 
     reg.type = "PredLine", pch=19, col = pupfish$Group)
legend("topleft", legend = unique(pupfish$Group), pch = 21, pt.bg = unique(pupfish$Group))
mtext("Common Slopes")
plot(fit.unique, type = "regression", predictor = pupfish$logSize, 
     reg.type = "PredLine", pch=19, col = pupfish$Group)
legend("topleft", legend = unique(pupfish$Group), pch = 21, pt.bg = unique(pupfish$Group))
mtext("Unique Slopes")
par(mfcol = c(1,1))
```

## **3: Size-Shape PCA**

One can also generate a plot of size-shape space (sensu Mitteroecker et al. 2004):

```{r eval=TRUE}
pc.plot <- plotAllometry(fit.unique, size = pupfish$logSize, logsz = FALSE, method = "size.shape", pch = 19, col = pupfish$Group)
legend("bottomleft", legend = unique(pupfish$Group), pch = 21, pt.bg = unique(pupfish$Group))
summary(pc.plot$size.shape.PCA)
```
