---
title: "Days 3 and 4:Disparity Analysis"
author: ""
output: html_document
---

## **Content Description**

In `geomorph`, the analysis of disparity is accomplished using Procrustes variance, which is one means of estimating multivariate variation in a dataset. PV is obtained for each group and compared using permutation. Let´s see some examples:

```{r eval=TRUE}
library(geomorph)
```

## Pupfish shape disparity

### Disparity around group means

First, although it is not needed, fitting a linear model is not a bad idea.

```{r eval=TRUE}
data("pupfish")
Group <- pupfish$Group <- interaction(pupfish$Pop, pupfish$Sex)
fit <- procD.lm(coords ~ Group, data = pupfish)
```

A simple way to perform a disparity analysis is with the pairwise function. 

```{r eval=TRUE}
PW <- pairwise(fit, groups = Group)
summary(PW, test.type = "var")
```

This approach is a little restricted in terms of options.  Another way to perform the analysis is by using the `geomorph::morphol.disparity` function.
```{r eval=TRUE}
MD <- morphol.disparity(fit, print.progress = FALSE)
summary(MD)
```

Although presented differently, the results are the same.  It is not necessary to have a linear model fit.  The formula for the linear model can be used directly in `morphol.disparity`.

```{r eval=FALSE, echo=TRUE}
MD <- morphol.disparity(coords ~ Group, group = Group,
                        data = pupfish,
                        print.progress = FALSE)
```

However, there is not a `plot.morphol.disparity` function, so having a linear model fit with `procD.lm` first offers some advantages.

```{r eval=FALSE, echo=TRUE}
P <- plot(fit, type = "PC", pch = 21, bg = pupfish$Group)
shapeHulls(P, Group, group.cols = c(1,3,2,4))
```

Feel free to also explore the shape space.

```{r, eval = FALSE}
picknplot.shape(P)
```

### Foote's (partial) Disparity

This requires using the `morphol.disparity` function rather than `pairwise`.

```{r eval=TRUE}
fit0 <- procD.lm(coords ~1, data = pupfish)
MD2 <- morphol.disparity(fit0, groups = Group,
                         partial = TRUE,
                         print.progress = FALSE)
summary(MD2)
```

The big difference is that each point is measured as a distance from the origin rather than the group mean.  A plot with hulls would not make sense (unless each hull also contained the origin).  The following plot is the best illustration of what is going on.

```{r}
P <- plot(fit, type = "PC")
PC <- P$PC.points[, 1:2]
for(i in 1 : nrow(PC)) 
  arrows(0, 0, PC[i,1], PC[i,2], col = pupfish$Group[i], length = 0.1)
```
