---
title: "05b. Disparity Analysis"
author: ""
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

## **Content Description**

In `geomorph`, the analysis of disparity is accomplished using Procrustes variance, which is one means of estimating multivariate variation in a dataset. PV is obtained for each group and compared using permutation. Let´s see some examples:

```{r eval=TRUE}
library(geomorph)
```

## **1: Salamanders**

```{r eval=TRUE}
data(plethodon)
Y.gpa <- gpagen(plethodon$land, print.progress = FALSE)    
gdf <- geomorph.data.frame(Y.gpa, species = plethodon$species, site = plethodon$site)

MD <- morphol.disparity(coords ~ species*site, groups = ~species*site, 
                  data = gdf, iter = 999, print.progress = FALSE)
summary(MD)
gp <- as.factor(paste(plethodon$species, plethodon$site))
plotTangentSpace(Y.gpa$coords, groups=gp)
```

One can look at variances of different groups than those modeled:

```{r eval=TRUE}
MD2 <- morphol.disparity(coords ~ species*site, groups = ~species, 
                         data = gdf, iter = 999, print.progress = FALSE)
summary(MD2)
```

One can also use Foote's definition of morphological disaprity (common mean)

```{r eval=TRUE}
MD3 <- morphol.disparity(coords ~ 1, groups = ~species, partial = TRUE,
                         data = gdf, iter = 999, print.progress = FALSE)
summary(MD3)
```

## **2: Pupfish**

```{r eval=TRUE}
data(pupfish) # data already aligned
gdf <- geomorph.data.frame(coords = pupfish$coords, 
                           CS = pupfish$CS,
                           Pop = pupfish$Pop,
                           Sex = pupfish$Sex)
MD <- morphol.disparity(coords ~ Pop*Sex, groups = ~ Pop*Sex, 
                        data = gdf, iter = 999, print.progress = FALSE)
summary(MD)
gp <- as.factor(paste(pupfish$Pop, pupfish$Sex))
plotTangentSpace(pupfish$coords, groups=gp)
```

The same approach, but considering variance associated with allometry.

```{r eval=TRUE}
pupfish.allometry <- procD.lm(coords ~ log(CS), ~Pop*Sex, data = gdf, iter = 999, print.progress = FALSE)
summary(pupfish.allometry)
MD2 <- morphol.disparity(pupfish.allometry, groups = gp, print.progress = FALSE)
summary(MD2)
```

An important point is that `morphol.disparity` can be used on a `procD.lm` fit.  Additionally, the `pairwise` function in `RRPP` can be used with a test type of "var" to compare variances.  Results are the same with either function.
