---
title: "Day-03. Shape Statistics II"
subtitle: "GLM, ANOVA, MANOVA, Model Comparisons"
author: ""
output: html_document
---

## **Content Description**

In `geomorph`, the main function for linear models. is `procD.lm`.  This module focuses less on most of the results `procD.lm` produces, but focuses more on model comparisons made available by `anova`, `maonova.update`, and `comprae.models`.

### Set up

```{r eval=TRUE}
library(geomorph)
data(pupfish) # GPA already performed
```

As a preliminary step, let's fit 10 different models with these data.

```{r}

fit0 <- procD.lm(coords ~ 1, 
                 data = pupfish, iter = 999)
fit1 <- procD.lm(coords ~ log(CS), 
                 data = pupfish, iter = 999)
fit2 <- procD.lm(coords ~ Sex, 
                 data = pupfish, iter = 999)
fit3 <- procD.lm(coords ~ Pop, 
                 data = pupfish, iter = 999)
fit4 <- procD.lm(coords ~ log(CS) + Sex, 
                 data = pupfish, iter = 999)
fit5 <- procD.lm(coords ~ log(CS) + Pop, 
                 data = pupfish, iter = 999)
fit6 <- procD.lm(coords ~ log(CS) * Sex, 
                 data = pupfish, iter = 999)
fit7 <- procD.lm(coords ~ log(CS) * Pop, 
                 data = pupfish, iter = 999)
fit8 <- procD.lm(coords ~ Sex * Pop, 
                 data = pupfish, iter = 999)
fit9 <- procD.lm(coords ~ log(CS) + Sex*Pop, 
                 data = pupfish, iter = 999)
fit10 <- procD.lm(coords ~ log(CS) * Sex*Pop, 
                  data = pupfish, iter = 999)

```

## Review model parts


Model that estimates a mean.  $\mathbf{X}$ is a vector of 1s and $\hat{\boldsymbol{\beta}}$ is a vector of mean values for each variable.

```{r}
# Model with just a mean

model.matrix(fit0)
coef(fit0)
```

Model that estimates an intercept and a slope.  $\mathbf{X}$ comprises a vector of 1s and a vector of $\log CS$.  The coefficients comprise a (row) vector of intercept values and a (row) vector of slopes.

```{r}
# Linear regression model

model.matrix(fit1)
coef(fit1)
```

Models that estimates group means.  $\mathbf{X}$ comprises a vector of 1s and a vector of a dummy variable.  The coefficients comprise a (row) vector of intercept values and a (row) vector of slopes.  However, the intercept is one group mean and the slopes are the changes from mean to another, by variable.

```{r}
# Simple single-factor models

model.matrix(fit2)
coef(fit2)
model.matrix(fit3)
coef(fit3)
```


## How fitted values are obtained.

Let's look at this in with just the first 4 variables, so it is easy to visualize

What the unique value of a model matrix look like:

```{r}
unique(model.matrix(fit2))
```

Multiply just these unique states by the coefficients.
```{r}
unique(model.matrix(fit2)) %*% coef(fit2)[,1:4]
```

Now see how that is replicated for every of $n$ observations

```{r}
model.matrix(fit2) %*% coef(fit2)[,1:4]
```

Should be the same as what `R` gives us if we just ask.

```{r}
fitted(fit2)[, 1:4]
```

# ANOVA

The following is a reminder of how ANOVA can do multiple model comparisons, using $R^2$ and $F$ as statistics for RRPP.

```{r}
anova(fit8, effect.type = "Rsq")
anova(fit8, effect.type = "F")
```

In case there is some confusion on the model comparisons performed.

```{r}
reveal.model.designs(fit8)
```


# MANOVA


The following is how MANOVA stats can be calculated, by updating the previous model.

```{r}
fitm <- manova.update(fit8, print.progress = FALSE)
```

Note that MANOVA statistics were simply appended to previous results.

```{r}
attributes(fit8)
attributes(fitm)
```

MANOVA summaries are provided with the `summary` function.

```{r}
summary(fitm)
summary(fitm, test = "Pillai")
```


# Pairwise comparisons, group means

Knowing means are different from ANOVA and MANOVA, it might be advisable to entertain which are different.

```{r}
group <- interaction(pupfish$Pop, pupfish$Sex)
PW <- pairwise(fit8, groups = group)

```

Two ways to view the same results:
```{r}
summary(PW)
summary(PW, stat.table = FALSE)
```

It is difficult to recommend how to use `summary.pairwise` until you get into the weeds with your own data.  `?pairwise` and `?summary.pairwise` will be helpful for finding ways to get into trouble, but getting out of trouble is elucidative for how to handle your data.

The next example shows how to compare slope vectors, which is more involved.

# Pairwise comparisons, group slopes

This example uses a model that allows for unique slopes among groups.  See if it makes sense how to estimate group-specific shape change  For example, which group would this vector, predict?  $\mathbf{x}^T = \begin{bmatrix} 1 & \log \bar{CS} &1 &0 & \log \bar{CS} & 0 & 0 & 0\\ \end{bmatrix}$

```{r}
coef(fit10)[, 1:2]
```

Notice below how we set up the pairwise object differently.

```{R, eval = FALSE}
group <- interaction(pupfish$Pop, pupfish$Sex)
PW <- pairwise(fit10, groups = group, 
               covariate = log(pupfish$CS))
```

We have to be very careful with the summary.  This summary probably does not make much sense:
```{r}
summary(PW)
```

These find the angle difference between the slopes of the groups.
```{r}
summary(PW, test.type = "VC")
summary(PW, test.type = "VC", angle.type = "deg")
```

These find the difference between the length of the slope vectors (difference in the amount of shape change).
```{r}
summary(PW, test.type = "DL")

```
