---
title: "Day-04: Asymmetry"
author: ""
output: html_document
---

## **Content Description**

Many biological objects display some sort of symmetry. Morphometric analyses often investigate symmetry in some manner: either by evaluating components of variation attributed to different symmetry effects (e.g., directional, fluctuating, etc.), or by accounting for symmetry prior to the analysis (e.g., obtain 'symmetrized' shapes for subsequent investigation).

```{r eval=TRUE}
library(geomorph)
```

## **Example of matching symmetry**

```{r eval=TRUE}
data(mosquito)
Y.gpa <- gpagen(mosquito$wingshape, print.progress = FALSE)
plot(Y.gpa)
mosquito.sym <- bilat.symmetry(A = Y.gpa, ind = mosquito$ind, side=mosquito$side, 
                           object.sym = FALSE, print.progress = FALSE)
summary(mosquito.sym)
```

## **Example of object symmetry**

```{r eval=TRUE}
data('lizards')
Y.gpa <- gpagen(lizards$coords, print.progress = FALSE)
plot(Y.gpa)

lizard.sym <- bilat.symmetry(A = Y.gpa, ind = lizards$ind, replicate = lizards$rep,
                           object.sym = TRUE, land.pairs = lizards$lm.pairs, print.progress = FALSE)
summary(lizard.sym)
plot(lizard.sym, warpgrids = TRUE)
```

*NOTE:* One may wish to remove asymmetry from the data and perform all subsequent analyses on 'symmetrized' structures. This can be easily achieved by obtaining the `$symm.shape` component from the `bilat.symmetry` object result.

### Scatterplots of specimens (full shape and symmetrized shape)
```{r eval = TRUE}
plotAllSpecimens(Y.gpa$coords)
plotAllSpecimens(lizard.sym$symm.shape)
```
