---
title: "04: Day 4 Material"
author: ""
output: html_document
---

```{r setup, include=FALSE, echo = TRUE, tidy = TRUE}
library(knitr)
library(geomorph)
opts_chunk$set(echo = TRUE)
```

## General Introduction

Today we will briefly introduce some `geomorph` functions for performing the analyses described yesterday. **Note:** You should be able to read in your data and perform the analyses from earlier this week! 

## 1: Trajectory Analysis

Here is a simple example of trajectory analysis: 

```{r eval=TRUE}
data(pupfish)
fit <- procD.lm(coords ~ Pop * Sex, 
                data = Pupfish, 
                iter = 999, 
                print.progress = FALSE)

reveal.model.designs(fit)

anova(fit)

TA <- trajectory.analysis(fit, 
                          groups = Pupfish$Pop, 
                          traj.pts = Pupfish$Sex)

summary(TA, attribute = "MD") # Magnitude difference (absolute difference between path distances)

summary(TA, attribute = "TC", angle.type = "deg") # Correlations (angles) between trajectories

# Plot results
TP <- plot(TA, 
           pch = as.numeric(Pupfish$Pop) + 20, 
           bg = as.numeric(Pupfish$Sex),
           cex = 0.7, col = "gray")

add.trajectories(TP, 
                 traj.pch = c(21, 22), 
                 start.bg = 1, 
                 end.bg = 2)

legend("topright", 
       levels(Pupfish$Pop), 
       pch =  c(21, 22), 
       pt.bg = 1)

```

## 2: Disparity Analysis

Here is a simple example of a disparity analysis: 

```{r eval=TRUE}
data("pupfish")

Group <- pupfish$Group <- interaction(pupfish$Pop, pupfish$Sex)

fit <- procD.lm(coords ~ Group, data = pupfish)

PW <- pairwise(fit, groups = Group)

summary(PW, test.type = "var")

# ALTERNATIVELY
MD <- morphol.disparity(fit, print.progress = FALSE)

summary(MD)

P <- plot(fit, type = "PC", pch = 21, bg = pupfish$Group)

shapeHulls(P, 
           Group, 
           group.cols = c(1, 3, 2, 4))
```

## 3: Integration and Modularity

Here are SOME of the integration/modularity analyses available: 

#### Overall Integration (and comparing it across regions)

```{r eval=TRUE}
data("plethodon")

Y.gpa <- gpagen(plethodon$land, 
                print.progress = FALSE)
  #Separate data by species

coords.gp <- coords.subset(Y.gpa$coords, 
                           plethodon$species)

#Z_Vrel by species
Vrel.gp <- Map(function(x) integration.Vrel(x), coords.gp) 

compare.ZVrel(Vrel.gp$Jord, Vrel.gp$Teyah)
```

#### Integration among subsets

```{r eval=TRUE}
data(pupfish) # GPA previously performed

group <- factor(paste(pupfish$Pop, pupfish$Sex, sep = "."))

# Subset 3D array by group, returning a list of 3D arrays
tail.LM <- c(1:3, 5:9, 18:38)
head.LM <- (1:56)[-tail.LM]
tail.coords <- pupfish$coords[tail.LM, , ]
head.coords <- pupfish$coords[head.LM, , ]

IT <- integration.test(tail.coords, 
                       head.coords, 
                       print.progress = FALSE)
summary(IT)
plot(IT)
```

#### Compare Strength of Integration acrouss groups

```{r eval=TRUE}
tail.coords.gp <- coords.subset(tail.coords, group)

head.coords.gp <- coords.subset(head.coords, group)

# Obtain Integration for groups
integ.tests <- Map(function(x,y) 
  integration.test(x, y, iter = 999, 
                print.progress = FALSE), 
  head.coords.gp, tail.coords.gp)

compare.pls(integ.tests)
```



#### Modularity

```{r eval=TRUE}
land.gp <- rep(1,56); land.gp[tail.LM] <- 2

MT <- modularity.test(pupfish$coords,
                      land.gp,
                      CI = FALSE,
                      print.progress = FALSE)

summary(MT)

plot(MT)
```

#### Compare Strength of Modularity Across Groups

```{r eval=TRUE}
coords.gp <- coords.subset(pupfish$coords, group)

modul.tests <- Map(function(x) 
  modularity.test(x, 
                  land.gp,
                  print.progress = FALSE), 
  coords.gp) 

compare.CR(modul.tests, CR.null = FALSE)
```


## 4: Data Prep/Clean-up

#### Estimating Missing Data

##### NOTE: First we *create* some missing data for illustration 
```{r eval=TRUE,out.width="70%"}
data(plethodon)

Y.gpa <- gpagen(plethodon$land, print.progress = FALSE)

plethland<-Y.gpa$coords
  plethland[3, , 2] <- plethland[8, , 2] <- NA  #create missing landmarks
  plethland[3, , 5] <- plethland[8, , 5] <- plethland[9,,5] <- NA  
  plethland[3, , 10] <- NA  
```

#### Now estimate 

```{r eval=TRUE,out.width="70%"}
new.tps <- estimate.missing(plethland, method = "TPS")
new.reg <- estimate.missing(plethland, method = "Reg")
```

##### Fix Positional Differences among regions

**See `fixed.angle`**

##### Combine landmark configurations

**See `combine.subsets`**

## 5: Things to Explore on Your Own

1. Remember Days 1 - 4
2. Perform phylogenetic ANOVA, phylogenetic PLS, and other PCMs. 
3. Visualize shapes from everything!
4. Plot summaries from these analyses to accompany the statistical findings
5. **Explore function options and output** (there are many!)

