---
title: "Day-04: Integration and Modularity"
author: ""
output: html_document
---

## **Content Description**

Characterizing patterns of integration and modularity are hugely important in morphometrics. At their core, these methods seek to summarize patterns of trait covariation, as found in the trait covariance matrix: $\hat{\mathbf\Sigma} = \mathbf{Z^T}\mathbf{Z}/ (n-1)$. There are three general classes of methods: 

  + 1: Methods for evaluating and comparing overall integration
  + 2: Methods for evaluating and comparing integration among subsets of traits
  + 3: Methods for evaluating and comparing modularity

`Geomorph implements approaches addressing all three of these (See lecture for detailed description of methods). 

```{r eval=TRUE}
library(geomorph)
```

## **1: Overall Integration**

Here we use an effect size based on $V_{rel}$ to quantify and compare overall levels of integration in a set of traits. 

```{r eval=TRUE}
data("plethodon")
Y.gpa <- gpagen(plethodon$land, print.progress = FALSE)
#Separate data by species
coords.gp <- coords.subset(Y.gpa$coords, plethodon$species)

#Z_Vrel by species
Vrel.gp <- Map(function(x) integration.Vrel(x), coords.gp) 
compare.ZVrel(Vrel.gp$Jord, Vrel.gp$Teyah)
```

## **1B: Overall Integration Across Spatial Scales**

A method for evaluating integration across spatial scales is also implemented

```{r eval=TRUE}
globalIntegration(Y.gpa$coords) #data are not spatially integrated
```

## **2: Integration between Subsets of Traits**

Here, one may use PLS to evaluate the covariation between subsets of traits: 


```{r eval=TRUE}
data(pupfish) # GPA previously performed
group <- factor(paste(pupfish$Pop, pupfish$Sex, sep = "."))

# Subset 3D array by group, returning a list of 3D arrays
tail.LM <- c(1:3, 5:9, 18:38)
head.LM <- (1:56)[-tail.LM]
tail.coords <- pupfish$coords[tail.LM,,]
head.coords <- pupfish$coords[head.LM,,]

IT <- integration.test(tail.coords, head.coords, print.progress = F)
summary(IT)
plot(IT)
```

One could also use a partition of landmarks (NOTE identical $r_{PLS}$: 

```{r eval=TRUE}
land.gp <- rep(1,56); land.gp[tail.LM] <- 2
integration.test(pupfish$coords, partition.gp=land.gp, print.progress = FALSE)
```

And the statistical results are the same as when using two.b.pls

```{r eval=TRUE}
two.b.pls(tail.coords, head.coords, print.progress = FALSE)
```

`picknplot.shape` example is provided in lab script.

## **2B: Comparing the Strength of Integration**

One may also evaluate whether the strength of integration similar in two or more groups. As above this is accomplished using effect sizes, this time based on $r_{PLS}$.

```{r eval=TRUE}
tail.coords.gp <- coords.subset(tail.coords, group)
head.coords.gp <- coords.subset(head.coords, group)

# Obtain Integration for groups
integ.tests <- Map(function(x,y) integration.test(x, y, iter=499, 
                print.progress = FALSE), head.coords.gp, tail.coords.gp)
compare.pls(integ.tests)
```

## **3: Modularity among Subsets of Traits**

Tests of modularity can also be performed in `geomorph`: 


```{r eval=TRUE}
MT <- modularity.test(pupfish$coords,land.gp,CI=FALSE,print.progress = FALSE)
summary(MT)
plot(MT)
```

## **3B: Comparing the Strength of Modularity**

One may also evaluate whether the strength of modularity similar in two or more groups. As above this is accomplished using effect sizes based on $CR$.

```{r eval=TRUE}
coords.gp <- coords.subset(pupfish$coords, group)
modul.tests <- Map(function(x) modularity.test(x, land.gp,print.progress = FALSE), coords.gp) 
compare.CR(modul.tests, CR.null = FALSE)
```

## **3B-2: Comparing Alternative Modular Hypotheses**

One may also wish to determine whether there is support in a given dataset for a 2-module, 3-module, 4-module, or other hypothesis. This is also accomplished via effect sizes of $CR$.

```{r eval=TRUE}
land.gps3 <- rep('a',56); land.gps3[39:48]<-'b'; land.gps3[c(6:9,28:38)] <- 'c' 
#3 module hypothesis (tail now a module)
land.gps4 <- rep('a',56); land.gps4[39:48]<-'b'; land.gps4[c(6:9,28:38)] <- 'c'; 
land.gps4[c(10,49:56)] <- 'd'  #4 module hypothesis (eye now a module)

m3.test <- modularity.test(coords.gp$Marsh.F,land.gps3, iter = 499, print.progress = FALSE)
m4.test <- modularity.test(coords.gp$Marsh.F,land.gps4, iter = 499, print.progress = FALSE)

model.Z <- compare.CR(m3.test,m4.test, CR.null = TRUE)
model.Z 
```
