---
title: "04.1: Asymmetry"
author: "Dean Adams, Michael Collyer & Antigoni Kaliontzopoulou"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

## **Content Description**

Many biological objects display some sort of symmetry. Morphometric analyses often investigate symmetry in some manner: either by evaluating components of variation attributed to different symmetry effects (e.g., directional, fluctuating, etc.), or by accounting for symmetry prior to the analysis (e.g., obtain 'symmetrized' shapes for subsequent investigation).

In `geomorph`, the analysis of (a)symmetry can be performed in two ways using the function `bilat.symmetry`. First, one can assess symmetry for 'matched' structures (e.g., left vs. right).  Alternatively, one can evaluate symmetry within a single structure (object symmetry).

```{r eval=TRUE}
library(geomorph)
```

## **Example of matching symmetry**

```{r eval=TRUE}
data(mosquito)
str(mosquito)
mosq.sym <- bilat.symmetry(mosquito$wingshape, ind = mosquito$ind, side = mosquito$side,
               replicate = mosquito$replicate, object.sym = FALSE, print.progress = FALSE)
summary(mosq.sym)
plot(mosq.sym)
```

## **Example of object symmetry**

```{r eval=TRUE}
data(scallops)
str(scallops)
plot(mshape(scallops$coorddata))
scallops$land.pairs

scallop.sym <- bilat.symmetry(scallops$coorddata, ind = scallops$ind, object.sym = TRUE,
                              land.pairs = scallops$land.pairs, print.progress = FALSE)
summary(scallop.sym)
plot(scallop.sym)
```

*NOTE:* One may wish to remove asymmetry from the data and perform all subsequent analyses on 'symmetrized' structures. This can be easily achieved by obtaining the `$symm.shape` component from the `bilat.symmetry` object result.


